### 2017-02-27 2017mergingmerges/section/03-Definition.tex
S3.2: talk about non-deterministic evaluation order, and lead into fusion

\subsection{Evaluation Order}

Consider the following example, where @alt2@ pulls two elements from its first input stream, then two from its second, before pushing all four to its output.
\begin{code}
  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
  alternates sIn1 sIn2 sIn3
   = let  sA1     = alt2 sIn1 sIn2
          sA2     = alt2 sIn2 sIn3
     in   zip sA1 sA2
\end{code}

Given the inputs @sIn1@ = $[a1, a2]$, @sIn2@ = $[b1, b2]$ and @sIn3@ = $[c1, c2]$ the \\ output will be $[ (a1, b1), (a2, b2), (b1, c1), (b2, c2) ]$

\TODO{Explain why this needs a specific evaluation order, and some do not work. Needs to fit on this page. Lead into the fusion order talk later.}

Although alt2 produces output elems two at a time, the consumer zip need its input elements to arrive alternately. At evaluation time we need the results pushed to sA1 and sA2 in the sA1 sA2 sA1 sA2 order, not sA1 sA1 sA2 sA2. Writing the rules nondeterministically allows the elaborator to discover a usable order, if there is one. This also affects fusion, we don't want to commit to the wrong order too early. We shall see that if we fuse the two alt processes first fusion will not work. We need to start with zip so that the order in which input elems arrive is constrained.

>>>>

% -----------------------------------------------------------------------------
\subsection{Non-deterministic Evaluation Order}

The evaluation rules of Fig. \ref{fig:Process:Eval:Feed} are non-deterministic in several ways. Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the group. More importantly, (ProcessPush) allows any process to perform a push action at any time, provided the other processes in the group are ready to accept the pushed value. Rule (FeedExternal) also allows new values pushed into an input channel from the environment, provided all processes that use that channel are ready to accept the value.

In our system, allowing the evaluation order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering. Consider the following example, where the @alt2@ operator pulls two elements form its first input stream, then two from the second, before pushing all four to its output.
\begin{code}
  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
  alternates sInA sInB sInC
   = let  s1   = alt2 sInA sInB
          s2   = alt2 sInB sInC
          sOut = zip s1 s2
     in   sOut
\end{code}

Note that the middle stream @sIn2@ is shared, and the result streams from both @alt2@ operators are zipped into tuples. Given the inputs @sInA@ = @[a1,a2]@, @sInB@ = @[b1,b2]@ and @sInC@ = @[c1,c2]@ the output of @zip@ will be @[(a1,b1),(a2,b2),(b1,c1),(b2,c2)]@, assuming @a1,a2,b1,b2@ and so on are values of type @Nat@.

Now, note that the first @alt2@ process pushes values to its output stream @s1@ two at a time, and the second @alt2@ process also pushes values to its own output stream @s2@ two at a time. However, the downstream @zip@ process needs to input one value from @s1@ then one from @s2@, then another from @s1@, then another from @s2@, alternating between the @s1@ and @s2@ streams. This will work, provided we can arrange for the two \emph{separate} @alt2@ processes to push to their separate output streams alternatively. They can still push two values at a time to their own outputs, but the downstream @zip@ process needs receive one from each process alternately. Here is a table of intermediate values to help make the explanation clearer:

\begin{code}
    sInA = [a1, a2, a3, a4, a5 ...]
    sInB = [b1, b2, b3, b4, b5 ...]
    sInC = [c1, c2, c3, c4, c5 ...]

    s1   = alt2 sInA sInB 
         = [a1, a2, b1, b2, a3, a4, b3, b4 ...]

    s2   = alt2 sInB sInC
         = [b1, b2, c1, c2, b3, b4, c3, c4 ...]

    sOut = zip s1 s2
         = [(a1,b1), (a2,b2), (b1,c1), (b2,c2) ...]
\end{code}

Considering the last line in the above table, note that @zip@ needs to output a tuple of @a1@ and @b1@ together, then @a2@ and @b2@ together, and so on. The implementation of the @zip@ process will attempt to pull the first value @a1@ from stream @s1@, blocking until it gets it, then pull the next value @b1@ from stream @s2@, blocking until it gets it. While @zip@ is blocked waiting for @b1@, the first @alt2@ process cannot yet push @a2@. The evaluation order of the overall process group is constrained by communication patterns of processes in that group.

As we cannot encode all possible evaluation orderings into the definition of the processes themselves, we have defined the evaluation rules to admit many possible orderings. In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, evaluation order would be discovered dynamically at runtime. In contrast, the role of our fusion transform is to determine one of these working evaluation orders statically. In the fused result process, the instructions will be scheduled so that they run in one of the orders that we would have gotten if the process group was executed dynamically. In doing so, we also eliminate the need to pass messages between processes --- once they are fused we can just copy values between heap locations.

% Although alt2 produces output elems two at a time, the consumer zip need its input elements to arrive alternately. At evaluation time we need the results pushed to sA1 and sA2 in the sA1 sA2 sA1 sA2 order, not sA1 sA1 sA2 sA2. Writing the rules nondeterministically allows the elaborator to discover a usable order, if there is one. This also affects fusion, we don't want to commit to the wrong order too early. We shall see that if we fuse the two alt processes first fusion will not work. We need to start with zip so that the order in which input elems arrive is constrained.
